Search techniques for gaussian mixture modelling with Minimum Message Length, with applications to the origin of our Galaxy

### 2018-04-20 article/ms.tex
foo

While we can evaluate the objective function $I(\vec{\data}|\vec{\psi})$, where 
for the sake of brevity we denote $\vec{\psi} = \{K, N, D, \vec{\weight}, 
\vecmean, \veccov\}$, and we can meaningfully compare models with different numbers of components or parameter
values, we require a search strategy in order to efficiently find the global
solution: the minimum message length over a class of finite mixture models
(where $N > K$). Before we introduce and compare search strategies,
let us first make some comments and quantitative descriptions on how the
total message length $I$ will vary with the model parameters. Specifically, we
will re-arrange Equation \ref{eq:objective_function} to be a sum of three 
smaller functions, and we will derive limits and predictions for those 
quantites. In an ideal scenario, those limits and predictions would be able
to provide us with an estimate of the message length for a Gaussian mixture
before computing the mixture (i.e., before performing expectation-maximization.
We begin by re-casting Equation \ref{eq:objective_function} as
\begin{eqnarray}
I(\vec{\data}|\vec{\psi}) = I_{1}(K, N, D, \vec{\weight}) 
                          + I_{2}(K, N, \veccov)
                          + I_{3}(\vec{\psi}) \quad ,
\end{eqnarray}
\noindent{}where:
\begin{eqnarray}
    I_{1} &=& \left(1-\frac{D}{2}\right)K\log{2} + \log{\Gamma(K)} + \frac{1}{4}\left[2\left(K - 1\right) + KD\left(D+3\right)\right]\log{N} \nonumber \\
          &+& \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \left(\frac{D(D+3) - 2}{4}\right)\sum_{k=1}^{K}\log{\weight_k} \todo{+ \kappa} \quad , \label{eq:I1} \\
    I_{2} &=& -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{|\veccov_k|} \quad {\textrm and},\\
    I_{3} &=& -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) - DN\log{\sigma_{\datum}} \quad .
\end{eqnarray}
The nomenclature for $I_1$, $I_2$ and $I_3$ bear no relation to the two-part 
message that we introduced earlier to describe message length. Rather we are
simply re-arranging the objective function because, as we will show, there are
useful theoretical bounds and predictions that we can make for each part of
this re-arranged expression.

@@

\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have 
derived the bounds
\begin{equation}
-K\log{K} \leq \sum_{k=1}^{K}\log\weight_k \leq - K\log{K} + \log(N + 1 - K)
\end{equation}
\noindent{}which gives us analytic expressions for the upper and lower value 
of $I_1$, regardless of the values of $\vec{\weight}$, for any mixture.

Developing similar theoretical bounds for $I_2$ is not straightforward, because
the determinants of the covariance matrices of Gaussian components will not
-- to our knowledge -- strictly follow some expected distribution, in the way 
that the weights $\weights$ follow a multinomial distribution. Similarly, the
\emph{scale} of $|\veccov_k|$ will depend entirely on the distribution of the
data. In that sense, we cannot a priori state with any certainty about what 
values $|\veccov_k|$ might take (or $\log{|\veccov_k|}$) without knowing
something about the data first. Even demanding that the covariance matrices
$\veccov_k$ are positive semi-definite only informs us that $\log{|\veccov_k|}$
is finite. However, there are quantitative predictions we can make on the 
bounds of $\sum_{k=1}^{K}\log{|\veccov_k|}$ if something is known about the
data. Specifically if we consider the determinant of the sample covariance 
matrix, which we denote $|\cov_{s}|$, then we can state with confidence that
the sample covariance matrix describes the maximum amount of variance that 
could be captured by the components in the mixture. For example, a limiting
case of $\sum_{k=1}^{K}\log{|\veccov_k|}$ can be described when the determinant
of the sample covariance matrix must be equally (uniformly) captured by $K$ 
components in a mixture. In other words, the maximum variance in the data 
must be equally spread between the components in the mixture. Uniformly 
describing the variance in the data is the limiting case because any 
non-uniformity would lead to smaller $|\veccov|$ for one mixture (and larger 
for another), and that asymmetry would lead to a higher $I_2$ because 
$\log(x - \epsilon) + \log(x + \epsilon) < 2\log{x}$, and because of the 
$-(1 + \frac{D}{2})$ term that enters multiplicatively in the calculation of 
$I_2$. Thus, we can specify a lower bound on $I_2$ based on the determinant of
the sample covariance matrix $|\cov_{s}|$.

The other extreme on $I_2$ can be approximated by estimating the \emph{minimum}
possible variance that could be captured by any single component.

>>>>

While we can evaluate the objective function $I(\vec{\data}|\vec{\psi})$, 
where for the sake of brevity we denote $\vec{\psi} = \{K, N, D, \vec{\weight}, 
\vecmean, \veccov\}$, and we can meaningfully compare models with different 
numbers of components or parameter values, we require a search strategy in 
order to efficiently find the global solution: the minimum message length 
over a class of finite mixture models (where $N > K$). Before we introduce and 
compare search strategies, let us first make some comments and quantitative 
descriptions on how the message length $I$ will vary with the model parameters. 
Specifically, we will re-arrange Equation \ref{eq:objective_function} to be a 
sum of three smaller functions, and we will derive limits and predictions for 
those quantites. In an ideal scenario, those limits and predictions would be 
able to provide us with an estimate of the message length for a Gaussian 
mixture before computing the mixture (i.e., before performing 
expectation-maximization.We begin by re-casting Eq. \ref{eq:objective_function} 
as
\begin{eqnarray}
I(\vec{\data}|\vec{\psi}) = I_{1}(K, N, D, \vec{\weight}) 
                          + I_{2}(K, N, \veccov)
                          + I_{3}(\vec{\psi}) \quad ,
\end{eqnarray}
\noindent{}where:
\begin{eqnarray}
    I_{1} &=& \left(1-\frac{D}{2}\right)K\log{2} + \log{\Gamma(K)} + \frac{1}{4}\left[2\left(K - 1\right) + KD\left(D+3\right)\right]\log{N} \nonumber \\
          &+& \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \left(\frac{D(D+3) - 2}{4}\right)\sum_{k=1}^{K}\log{\weight_k} \todo{+ \kappa} \quad , \label{eq:I1} \\
    I_{2} &=& -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{|\veccov_k|} \quad {\textrm and},\\
    I_{3} &=& -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) - DN\log{\sigma_{\datum}} \quad .
\end{eqnarray}
The nomenclature for $I_1$, $I_2$ and $I_3$ bear no relation to the two-part 
message that we introduced earlier to describe message length. Rather we are
simply re-arranging the objective function because, as we will show, there are
useful theoretical bounds and predictions that we can make for each part of
this re-arranged expression.

@@

\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have 
derived the bounds
\begin{equation}
-K\log{K} \quad \leq \quad \sum_{k=1}^{K}\log\weight_k \quad \leq \quad - K\log{K} + \log(N + 1 - K)
\end{equation}
\noindent{}which gives us analytic expressions for the upper and lower value 
of $I_1$, regardless of the values of $\vec{\weight}$, for any mixture, and
bounds on $I_{1}$ can be derived from just $\{K, N, D\}$.

Developing similar theoretical bounds for $I_2$ is not straightforward, because
the determinants of the covariance matrices of Gaussian components will not
-- to our knowledge -- strictly follow some expected distribution, in the way 
that the weights $\weights$ follow a multinomial distribution. Similarly, the
\emph{scale} of $|\veccov_k|$ will depend entirely on the distribution of the
data. In that sense, we cannot a priori state with any certainty about what 
values $|\veccov_k|$ might take (or $\log{|\veccov_k|}$) without knowing
something about the data first. Even demanding that the covariance matrices
$\veccov_k$ are positive semi-definite only informs us that $\log{|\veccov_k|}$
is finite. However, there are quantitative predictions we can make on the 
bounds of $\sum_{k=1}^{K}\log{|\veccov_k|}$ if something is known about the
data. Specifically if we consider the determinant of the sample covariance 
matrix, which we denote $|\cov_{u}|$ (where the $u$ subscript denotes `upper',
for the upper limit on the covariance matrix), then we can state with confidence that
the sample covariance matrix describes the maximum amount of variance that 
could be captured by the components in the mixture. For example, a limiting
case of $\sum_{k=1}^{K}\log{|\veccov_k|}$ can be described when the determinant
of the sample covariance matrix must be equally (uniformly) captured by $K$ 
components in a mixture. In other words, the maximum variance in the data 
must be equally spread between the components in the mixture. Uniformly 
describing the variance in the data is the limiting case because any 
non-uniformity would lead to smaller $|\veccov|$ for one mixture (and larger 
for another), and that asymmetry would lead to a higher $I_2$ because 
$\log(x - \epsilon) + \log(x + \epsilon) < 2\log{x}$, and because of the 
$-(1 + \frac{D}{2})$ term that enters multiplicatively in the calculation of 
$I_2$. Thus, we can specify a lower bound on $I_2$ based on the determinant of
the sample covariance matrix $|\cov_{u}|$:

\begin{eqnarray}
    I_{2} & \geq & -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{\frac{|\cov_u|}{K}} \nonumber \\
    I_{2} & \geq & -\left(1 + \frac{D}{2}\right)\left(K\log{|\cov_u|} - K\log{K}\right) 
\end{eqnarray}

The other extreme on $I_2$ can be approximated by estimating the \emph{minimum}
possible variance that could be captured by any single component. In theory
this approaches zero, but in practice it will be set by the minimum pairwise
distance in $D$ dimensions. Specifically, if we consider two data points to be
captured by a component in the mixture, an those two data points are the
closest two data points (in high dimensional space) of any data points, then
the variance captured by that Gaussian component will represent the \emph{minimum}
possible variance that can be captured by any single component.  
In practice we calculate the minimum total pair-wise euclidean distance $d_{min}$
in $D$ dimensions, average it over the dimensions, and approximate the minimum
determinant of the covariance matrix $|\cov_l|$ as
\begin{equation}
    |\cov_{l}| \approx \prod_{d=1}^{D}\frac{d_{min}^2}{4} \quad ,
\end{equation}
\noindent{}which approximately represents the variance between two data points,
but ignores correlations and assumes only a mean distance in $D$ dimensions
(for comuptational cost). With these approximations,
\begin{eqnarray}
    \log{|\cov_{l}|} & \approx & \sum_{d=1}^{D}\log{\frac{d_{min}^2}{4}}  \nonumber \\
    \log{|\cov_{l}|} & \approx & 2D\log{\left(\frac{d_{min}}{2}\right)} \quad ,
\end{eqnarray}
\noindent{}and the upper limit on $I_2$ is:
\begin{eqnarray}
    I_{2} & \leq & -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{|\cov_l|} \nonumber \\
    I_{2} & \leq & -DK\left(D+2\right)\log{\left(\frac{d_{min}}{2}\right)} \quad .
\end{eqnarray}

Here our bounds on $I_2$ are only dependent on $K$, $D$, the minimum euclidean
pair-wise distance $d_{min}$, and the determinant of the sample covariance
matrix $|\cov_u|$. In practice it is the lower bound on $I_{2}$ that is 
arguably more useful for search and optimisation, but one could conceive
situations where there is great utility in also having upper bounds on $I_1$
and $I_2$.

There is no finite upper bound on $I_{3}$ because the negative log likelihood
$-\log\likelihood$ will approach infinity for an increasingly poor fit to the
data. A lower limit, however, does exist.

\andy{here}
